Characterizing the Splice Map of Turkey Hemorrhagic Enteritis Virus

### 2024-07-09 manuscript_thev_transcriptome.tex
addressed reviewer comments from virology journal

\textbf{Background:} Hemorrhagic enteritis, caused by \emph{Turkey
Hemorrhagic Enteritis Virus (THEV)}, is a disease affecting turkey
poults characterized by immunosuppression and bloody diarrhea. An
avirulent THEV strain that retains the immunosuppressive ability is used
as a live vaccine. Characterizing the splice map of THEV is an essential
step that would allow studies of individual genes mediating its
immunosuppressive functions. We used RNA sequencing to characterize the
splice map of THEV for the first time, providing key insights into the
THEV gene expression and mRNA structures.\\
\textbf{Methods:} After infecting a turkey B-cell line with the vaccine
strain, samples in triplicates were collected at 4-, 12-, 24-, and
72-hours post-infection. Total RNA was extracted and poly-A-tailed mRNA
sequenced. Reads were mapped to the THEV genome after trimming and
transcripts assembled with StringTie. We performed PCR of THEV cDNA,
cloned the PCR products, and used Sanger sequencing to validate all
identified splice junctions.\\
\textbf{Results:} Researchers previously annotated the THEV genome as
encoding 23 open reading frames (ORFs). We identified 29 spliced
transcripts from our RNA sequencing data, all containing novel exons
although some exons matched some previously annotated ORFs. The three
annotated splice junctions were also corroborated by our data. During
validation we identified five additional unique transcripts, a subset of
which were further validated by 3' rapid amplification of cDNA ends (3'
RACE). Thus, we report that the genome of THEV contains 34 transcripts
with the coding capacity for all annotated ORFs. However, we found six
of the previously annotated ORFs to be truncated ORFs on the basis of
the identification of an in-frame upstream start codon or the detection
of additional coding exons. We also identified three of the annotated
ORFs with longer or shorter isoforms, and seven novel unannotated ORFs
that could potentially be translated; although it is beyond the scope of
this manuscript to investigate whether they are translated.\\
\textbf{Conclusions:} Similar to human adenoviruses, all THEV
transcripts are spliced and organized into five transcription units
under the control of their cognate promoters. The genes are expressed
under temporal regulation and THEV also produces multiple distinctly
spliced transcripts that code for the same protein.

@@

Adenoviruses (AdVs) are non-enveloped icosahedral-shaped DNA viruses,
causing infection in virtually all types of vertebrates studied to date.
Their double-stranded linear DNA genomes range between 26 and 45kb in
size, producing a broad repertoire of transcripts via highly complex
alternative splicing patterns (1, 2). The AdV genome is one of the most
optimally economized; both the forward and reverse DNA strands harbor
protein-coding genes, making it highly gene-dense. There are 16 genes
termed ``genus-common'' that are homologous in all AdVs, presumably
inherited from a common ancestor. All other genes are termed
``genus-specific''. The genus-specific genes tend to be located at the
termini of the genome while genus-common genes are usually towards the
center of the genome (1). This pattern is also observed in
\emph{Poxviridae} and \emph{Herpesviridae}, which also have linear DNA
genomes (1, 3, 4). The family \emph{Adenoviridae} consists of five
genera: \emph{Mastadenovirus} (MAdV), \emph{Aviadenovirus},
\emph{Atadenovirus}, \emph{Ichtadenovirus}, and \emph{Siadenovirus}
(SiAdV) to which turkey adenovirus 3 also called turkey hemorrhagic
enteritis virus (THEV) belongs (5--10). Members of SiAdV have the
smallest genome size (\textasciitilde26 kb) and gene content of all
known AdVs (see \textbf{Figure 1}) (1, 2, 6).

Virulent THEV strains (THEV-V) and avirulent strains (THEV-A) of THEV
both infect turkeys, with THEV-V causing hemorrhagic enteritis (HE), a
debilitating acute disease predominantly affecting turkey poults
characterized by immunosuppression, intestinal lesions leading to bloody
diarrhea, and up to 80\% mortality (2, 11--13). While the current
vaccine strain (a THEV-A called Virginia Avirulent Strain {[}VAS{]}) has
proven effective at preventing HE in turkey poults, it still retains its
immunosuppressive ability. Thus, vaccinated birds are rendered more
susceptible to opportunistic infections and death than unvaccinated
birds leading to substantial economic losses (11, 14--16). To eliminate
the immunosupressive effect of the vaccine strain, a thorough
investigation of the culprit viral genes mediating this phenomenon is
essential. However, the transcriptome (splicing and gene expression
patterns) of THEV has not been characterized, making an investigation of
specific immunosuppressive viral genes impractical.

A myriad of studies have elucidated the AdV transcriptome in fine detail
(17, 18). However, a large preponderance of studies focus on MAdVs --
specifically human AdVs. Thus, most of the current AdV gene expression
and replication knowledge is based on MAdV studies, which is generalized
for all other AdVs (10, 19). MAdV transcription is temporally regulated;
their genes are categorized into five early transcription units (E1A,
E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and
one major late transcription unit (MLTU or major late promoter {[}MLP{]}
region), which generates five families of late mRNAs (L1-L5) based on
the polyadenylation site. An additional gene (UXP or U exon) is located
on the reverse strand. The early genes encode non-structural proteins
such as enzymes or host-cell modulating proteins, primarily involved in
DNA replication, or providing the necessary intracellular niche for
optimal replication while late genes encode structural proteins that act
as capsid proteins, promote virion assembly, or direct genome packaging.
The immediate early genes E1A are expressed first, followed by the
delayed early genes, E1B, E2, E3 and E4. Then the intermediate early
genes, IVa2 and pIX are expressed followed by the late genes (10, 17,
18). It is noteworthy that the MLP shows basal transcriptional activity
during early infection (before DNA replication), with a comparable
efficiency to other early viral promoters, but it reaches its maximal
activity during late infection (after DNA replication). However, during
early infection only a subset of the MLP-derived transcripts are
expressed (10). MAdV makes an extensive use of alternative RNA splicing
to produce a very complex array of mRNAs. All but the pIX mRNA undergo
at least one splicing event. For instance, the MLTU produces over 20
distinct splice variants all containing three non-coding exons at the
5'-end (collectively known as the tripartite leader; TPL) (17, 18).
There is also an alternate three-exon 5' non-coding leader sequence
present in varying amounts on a subset of MLTU mRNAs (known as the x-,
y-, and z-leaders). Lastly, there is the i-leader exon, which is
infrequently included between the second and third TPL exons, and codes
for the i-leader protein (20). Thus, the MLTU produces a complex
repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced
onto different 3' coding exons which are grouped into five different
3'-end classes (L1-L5) based on polyadenylation site. Each transcription
unit (TU) contains its own promoter driving the expression of the array
of mRNA transcripts produced via alternative splicing in the unit (10,
17, 18). The promoters are activated at different phases of infection by
proteins from previously activated TUs. Paradoxically, the early-to-late
phase transition during infection requires the L4 gene products, 22K and
33K, which should only be available after the transition. However, a
promoter in the L4 region (L4P) that directs the expression of these two
proteins independent of the MLP was found, resolving the paradox (10,
17, 21). During translation of AdV mRNA, recent studies using long-read
direct RNA sequencing strongly suggest the potential usage of secondary
start codons; adding to what was already a highly complex system for
gene expression (17, 22).

@@

\textbf{Overview of sequencing data and analysis pipeline outputs}\\
A prior study by Aboezz \emph{et al}. demonstrated that nearly all THEV
transcripts became detectable starting at 4 hours post-infection (hpi),
with one replication cycle concluding around 18 hpi (24). Consequently,
we harvested infected MDTC-RP19 cells (MOI of 100 genome copy
numbers/cell) at 4-, 12-, 24-, and 72-hpi to capture all transcripts
within a broad time window. Our paired-end RNA sequencing (RNA-seq)
experiment generated an average of 107.1 million total reads of 149 bp
length per time-point. These reads were concurrently mapped to both the
virus (THEV) and host (\emph{Meleagris gallopavo}) genomes using the
Hisat2 (25) reference-based aligner. A total of 18.1 million reads from
all time-points mapped to the virus genome, providing comprehensive
coverage and leaving no regions unmapped. The mapped reads to the virus
genome increased significantly from a scant 432 reads at 4 hpi to 16.9
million reads at 72 hpi (\textbf{Table 1}, \textbf{Figure 2A}). From
these mapped reads, we identified 2,457 unique THEV splice junctions
across all time-points, with later time-points exhibiting significantly
more sequence reads supporting the splice junctions than earlier
time-points. For instance, all 13 unique junctions at 4 hpi had fewer
than 10 supporting reads each, averaging only 2.8 reads per junction. In
contrast, the 2374 unique junctions at 72 hpi averaged 898.4 reads per
junction, with some junctions reaching as high as 322,677 reads. The
marked increase in splice junction and mapping reads to the THEV genome
over time indicates an active infection and successful viral
replication, which is corroborated by our quantitative PCR (qPCR) assay
that quantified the total number of viral genome copies over time
(\textbf{Figure 2B}).

@@

The E2B region transcripts also start with the E2-5'UTR but extend
downstream to reach the TTS at 2334 bp in the IM region, which is in the
immediate context of a canonical PAS (position 2333 bp) where
polyadenylation likely occurs. The E2B transcripts, TRXPT\_6 and
TRXPT\_7, are almost identical except for an extra splice junction at
the 3'-end of TRXPT\_6 (\textbf{Figure 7}). TRXPT\_7 has the CP for both
classical proteins (pTP and Ad-pol) encoded in this region, with the pTP
ORF predicted to be spliced from two exons. The predicted splice
junction of pTP is corroborated by our data but the full transcript is
markedly longer than the predicted ORF, although the encoded product
(pTP) remains unchanged. Ad-pol (polymerase) is encoded downstream of
pTP with secondary start codon (secSC) usage. The CP of TRXPT\_6
slightly differs from TRXPT\_7 because a new STC resulting from the
extra splice site forms a minimal truncation of the Ad-pol encoded from
its secSC.

@@

\textbf{Major Late Transcription Unit (MLTU) or MLP Region
transcripts}\\
The MLTU transcripts, dominant in the late phase of the AdV infectious
cycle, are produced by alternative polyadenylation and splicing of a
primary transcript and grouped into five transcript classes (L1-L5).
About 13 out of the 23 predicted ORFs in THEV fall within this TU, some
of which we have categorized under the E3 TU instead. Our RNA-seq data
revealed 12 transcripts (TRXPT\_8, TRXPT\_9, TRXPT\_10, TRXPT\_11,
TRXPT\_12, TRXPT\_13, TRXPT\_14, TRXPT\_16, TRXPT\_17, TRXPT\_18,
TRXPT\_19, TRXPT\_20) in this TU, most of which have the 5' TPL sequence
as in all AdVs. However, three transcripts (TRXPT\_16, TRXPT\_17,
TRXPT\_18) use a different leader sequence (sTPL), where a different
first exon is used instead of the first TPL exon, and TRXPT\_20 uses
only the third TPL exon (TPL3); see \textbf{Figure 10}.

@@

While the advent of next-generation sequencing has rendered easier the
study of large and complex eukaryotic transcriptomes, studying the
smaller, compact viral transcriptomes is counterintuitively more
challenging, as the transcripts typically have significant overlaps due
to genome economization. AdV transcriptomes escalate the difficulty due
to the wide array of mRNAs produced via very complex alternative
splicing and polyadenylation, all initiated from relatively few
promoters. Standard RNA-seq analysis programs, not primarily designed
for such compact, gene-dense, and complex transcriptomes, further
compound this challenge. Furthermore, in our case, no prior
transcriptomic studies for THEV exist. Our approach combines standard
RNA-seq analysis programs with custom analyses and experimentally
validating all splice junctions with independent methods. The transcript
map for THEV produced from our analysis is strikingly similar to that of
the MAdVs.

Our work provides the first insights into THEV splicing, revealing 34
transcripts grouped into five transcription units (TUs). The general
temporal gene expression regulation observed in MAdVs, with early
regions peaking at earlier time points followed by MLTU predominance at
later time points, seems to also apply to THEV. An unexpected
observation is that the pileup of mapped reads to THEV seems
consistently skewed over similar regions of the genome at all time
points. Given the temporal regulation of AdVs gene expression, we
anticipated distinct differences in read pileups over the genome at
different time points, indicating the different stages of infection.
This may be due to an unsynchronized infection, leading to transcripts
overlapping the time points. Further research is required to determine
the precise temporal regulation of THEV.

Short read deep sequencing effectively reconstructs full AdV mRNA
structures, particularly mapping splice sites (18). However, the
substantial overlapping nature of AdV mRNAs and fragmentation during
library preparation make it challenging to map the exact TSS, TTS, and
PASs of assembled transcripts. Also, transcripts with significant
overlaps may be assembled as a single, longer mRNA, since the short
reads alone do not provide enough context for the transcript assembler
(StringTie) to distinguish them. Our results show transcripts in the
same TU initiating or terminating in similar areas, but not at the exact
same position. We consider the most upstream TSS or most downstream TTS
for for the transcripts involved but we present them unchanged in all
the figures shown (see \textbf{Supplementary Table S2}). Also, comparing
our results to the better-studied MAdV transcriptomes, we believe some
long transcripts in the MLTU (TRXPT\_9, TRXPT\_10, and TRXPT\_11) are
likely due to fusing some E3 transcripts to the terminal exons of the
MLTU transcripts by StringTie, making them significantly longer. These
mRNAs have unusually many exons and their last few exons are identical
to some E3 mRNAs. Future studies using long read sequencing technologies
will provide more precise mapping of the TSS and TTS and clarify the
structures of the long MLTU transcripts.

While most predicted ORFs are encoded by the spliced transcripts, we
found some that seem to be truncated predictions, as either an upstream
in-frame SC or unknown upstream exons were found. Other ORFs were
identified that were either shorter or longer isoforms of some predicted
ORF. We also found several novel unpredicted ORFs (\textbf{Supplementary
Table S3}). On this basis, we anticipate that further studies will
likely reveal more unpredicted novel ORFs or new variants of predicted
ORFs. Furthermore, it is not unreasonable to presume that several splice
variants will likely be found as evidenced firstly by finding unique
transcripts using 3' RACE and during our splice junction validation
steps. And secondly, recent studies (17, 18, 22) are still discovering
novel mRNA variants for even the best studied MAdVs decades later.

Eukaryotic mRNAs are typically functionally monocistronic, with the
5'-most AUG determining the translation reading frame. However, AdV
mRNAs, which span more than one ORF, are functionally polycistronic,
employing non-standard mechanisms of translation initiation such as
secSC usage and ribosome shunting (10, 22). AdVs use secondary AUGs as
initiation codons for most E1b proteins and some E3 proteins. In fact,
recent studies show that secSC usage is found transcriptome-wide. This
is thought to occur because translation initiation at the first SC is
inefficient, allowing downstream SCs to be employed (17). Ribosomal
shunting or jumping mechanism is utilized for MLTU transcripts that have
the TPL. This mechanism allows the ribosome to translocate to a
downstream AUG, under the direction of the shunting elements in the TPL,
even if a start codon in a good Kozak sequence context is bypassed.
Thus, predicting the protein(s) expressed from an AdV mRNA is uncertain
as any one of the AUGs may be selected (10, 22). Almost all the THEV
transcripts in our data have the CP for several ORFs, some spanning as
many as six ORFs. This supports the usage of these special ribosome
initiation mechanisms as several predicted and novel ORFs found on mRNA
in our data could not be translated using only the typical ribosome
scanning mechanism. Interestingly, several distinct THEV mRNAs have
identical CPs. This is also observed in human AdVs in a recent study
(17). They proposed that this may permit protein production to be
fine-tuned through alteration in the balance between different mRNA
groups expressing that ORF.

@@

The THEV transcriptome bears remarkable similarity to the better-studied
MAdVs. The transcriptome is organized into five TUs, with temporal
regulation divided into early and late genes, and a broad repertoire of
transcripts are produced via virtually unlimited alternative splicing.
However, the THEV transcriptome appears less sophisticated (i.e, it
encodes fewer genes) than MAdVs, primarily because the MAdV genomes are
close to twice as long as that of THEV. The lack of subdivision of the
E1 region into E1a and E1b is one of the most obvious examples. Also,
the MAdV E4 region encodes several proteins unlike in THEV where only
one transcript encoding one protein was found. The complexity of the
MLTU leader sequences is another example. While the majority of the THEV
MLTU transcripts begin with the TPL just like MAdVs with a small subset
using a variant leader sequence (sTPL), significantly more diverse
5'-UTRs are employed for MAdV MLTU transcripts. Namely, the TPL, the
so-called x, y, and z leaders, and the i-leader are 5' leaders utilized
by MAdV MLTU mRNAs. The absence of these non-TPL leaders in our data
could mean that the 5'-UTR diversity of THEV's MLTU mRNAs is more
limited due to its smaller genome size or future studies could uncover
more variety not seen in our results. Being the first transcriptomic
characterization of THEV, this work should serve as useful resource to
future THEV gene expression and transcriptomic studies, especially,
mapping the mRNA splice sites. \newpage

>>>>

\textbf{Background:} Hemorrhagic enteritis, caused by \emph{Turkey
Hemorrhagic Enteritis Virus (THEV)}, is a disease affecting turkey
poults characterized by immunosuppression and bloody diarrhea. An
avirulent THEV strain that retains the immunosuppressive ability is used
as a live vaccine. Characterizing the splice map of THEV is an essential
step that would allow studies of individual genes mediating its
immunosuppressive functions. We used RNA sequencing to characterize the
splice map of THEV for the first time, providing key insights into the
THEV gene expression and mRNA structures.\\
\textbf{Methods:} After infecting a turkey B-cell line with the vaccine
strain, samples in triplicates were collected at 4-, 12-, 24-, and
72-hours post-infection. Total RNA was extracted, and poly-A-tailed mRNA
sequenced. Reads were mapped to the THEV genome after trimming and
transcripts assembled with StringTie. We performed PCR of THEV cDNA,
cloned the PCR products, and used Sanger sequencing to validate all
identified splice junctions.\\
\textbf{Results:} Researchers previously annotated the THEV genome as
encoding 23 open reading frames (ORFs). We identified 29 spliced
transcripts from our RNA sequencing data, all containing novel exons
although some exons matched some previously annotated ORFs. The three
annotated splice junctions were also corroborated by our data. During
validation we identified five additional unique transcripts, a subset of
which were further validated by 3' rapid amplification of cDNA ends (3'
RACE). Thus, we report that the genome of THEV contains 34 transcripts
with the coding capacity for all annotated ORFs. However, we found six
of the previously annotated ORFs to be truncated ORFs on the basis of
the identification of an in-frame upstream start codon or the detection
of additional coding exons. We also identified three of the annotated
ORFs with longer or shorter isoforms, and seven novel unannotated ORFs
that could potentially be translated; although it is beyond the scope of
this manuscript to investigate whether they are translated.\\
\textbf{Conclusions:} Similar to human adenoviruses, all THEV
transcripts are spliced and organized into five transcription units
under the control of their cognate promoters. The genes are expressed
under temporal regulation and THEV also produces multiple distinctly
spliced transcripts that code for the same protein. Studies of the newly
identified potential proteins should be urgently performed as these
proteins may have roles in THEV-induced immunosuppression. Also, knowing
the splicing of THEV genes should be invaluable to future research
focusing of studying THEV genes, as this will allow accurate cloning of
the mRNAs.

@@

Adenoviruses (AdVs) are non-enveloped icosahedral-shaped DNA viruses,
causing infection in virtually all types of vertebrates studied to date.
Their double-stranded linear DNA genomes range between 26 and 45kb in
size, producing a broad repertoire of transcripts via highly complex
alternative splicing patterns (1, 2). The AdV genome is one of the most
optimally economized; both the forward and reverse DNA strands harbor
protein-coding genes, making it highly gene-dense. There are 16 genes
termed ``genus-common'' that are homologous in all AdVs, presumably
inherited from a common ancestor. All other genes are termed
``genus-specific''. The genus-specific genes tend to be located at the
termini of the genome while genus-common genes are usually towards the
center of the genome (1). This pattern is also observed in
\emph{Poxviridae} and \emph{Herpesviridae}, which also have linear DNA
genomes (1, 3, 4). The family \emph{Adenoviridae} consists of five
genera: \emph{Mastadenovirus} (MAdV), \emph{Aviadenovirus},
\emph{Atadenovirus}, \emph{Ichtadenovirus}, and \emph{Siadenovirus}
(SiAdV) to which turkey adenovirus 3 also called turkey hemorrhagic
enteritis virus (THEV) belongs (5--10). Members of SiAdV have the
smallest genome size (\textasciitilde26 kb) and gene content of all
known AdVs, with five genus-specific genes of undefined functions (see
\textbf{Figure 1}) (1, 2, 6).

Virulent THEV strains (THEV-V) and avirulent strains (THEV-A) of THEV
both infect turkeys, with THEV-V causing hemorrhagic enteritis (HE), a
debilitating acute disease predominantly affecting turkey poults
characterized by immunosuppression, intestinal lesions leading to bloody
diarrhea, and up to 80\% mortality (2, 11--13). While the current
vaccine strain (a THEV-A called Virginia Avirulent Strain {[}VAS{]}) has
proven effective at preventing HE in turkey poults, it still retains its
immunosuppressive ability. Thus, vaccinated birds are rendered more
susceptible to opportunistic infections and death than unvaccinated
birds leading to substantial economic losses (11, 14--16). To eliminate
the immunosuppressive effect of the vaccine strain, a thorough
investigation of the culprit viral genes mediating this phenomenon is
essential. However, the transcriptome (splicing and gene expression
patterns) of THEV has not been characterized, making an investigation of
specific immunosuppressive viral genes impractical.

A myriad of studies have elucidated the AdV transcriptome in fine detail
(17, 18). However, a large preponderance of studies focus on MAdVs --
specifically human AdVs. Thus, most of the current AdV gene expression
and replication knowledge is based on MAdV studies, which is generalized
for all other AdVs (10, 19). MAdV transcription is temporally regulated;
their genes are categorized into five early transcription units (E1A,
E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and
one major late transcription unit (MLTU or major late promoter {[}MLP{]}
region), which generates five families of late mRNAs (L1-L5) based on
the polyadenylation site. An additional gene (UXP or U exon) is located
on the reverse strand. The early genes encode non-structural proteins
such as enzymes or host-cell modulating proteins, primarily involved in
DNA replication, or providing the necessary intracellular niche for
optimal replication while late genes encode structural proteins that act
as capsid proteins, promote virion assembly, or direct genome packaging.
The immediate early genes E1A are expressed first, followed by the
delayed early genes, E1B, E2, E3 and E4. Then the intermediate early
genes, IVa2 and pIX are expressed followed by the late genes (10, 17,
18). It is noteworthy that the MLP shows basal transcriptional activity
during early infection (before DNA replication), with a comparable
efficiency to other early viral promoters, but it reaches its maximal
activity during late infection (after DNA replication). However, during
early infection only a subset of the MLP-derived transcripts are
expressed (10). MAdVs make extensive use of alternative RNA splicing and
polyadenylation to produce a very complex array of mRNAs. All but the
pIX mRNA undergo at least one splicing event. For instance, the MLTU
produces over 20 distinct splice variants all containing three
non-coding exons at the 5'-end (collectively known as the tripartite
leader; TPL) (17, 18). There is also an alternate three-exon 5'
non-coding leader sequence present in varying amounts on a subset of
MLTU mRNAs (known as the x-, y-, and z-leaders). Lastly, there is the
i-leader exon, which is infrequently included between the second and
third TPL exons, and codes for the i-leader protein (20). Thus, the MLTU
produces a complex repertoire of mRNA with diverse 5' untranslated
regions (UTRs) spliced onto different 3' coding exons which are grouped
into five different 3'-end classes (L1-L5) based on polyadenylation
site. Each transcription unit (TU) contains its own promoter driving the
expression of the array of mRNA transcripts produced via alternative
splicing in the unit (10, 17, 18). The promoters are activated at
different phases of infection by proteins from previously activated TUs.
Paradoxically, the early-to-late phase transition during infection
requires the L4 gene products, 22K and 33K, which should only be
available after the transition. However, a promoter in the L4 region
(L4P) that directs the expression of these two proteins independent of
the MLP was found, resolving the paradox (10, 17, 21). During
translation of AdV mRNA, recent studies using long-read direct RNA
sequencing strongly suggest the potential usage of secondary start
codons; adding to what was already a highly complex system for gene
expression (17, 22).

@@

\textbf{Overview of sequencing data and analysis pipeline outputs}\\
A prior study by Aboezz \emph{et al}. demonstrated that nearly all THEV
transcripts became detectable starting at 4 hours post-infection (hpi),
with one replication cycle concluding around 18 hpi (24). Consequently,
we harvested infected MDTC-RP19 cells (multiplicity of infection (MOI)
of 100 genome copy numbers/cell) at 4-, 12-, 24-, and 72-hpi to capture
all transcripts within a broad time window. Our paired-end RNA
sequencing (RNA-seq) experiment generated an average of 107.1 million
total reads of 149 bp length per time-point. These reads were
concurrently mapped to both the virus (THEV) and host (\emph{Meleagris
gallopavo}) genomes using the Hisat2 (25) reference-based aligner. A
total of 18.1 million reads from all time-points mapped to the virus
genome, providing comprehensive coverage and leaving no regions
unmapped. The mapped reads to the virus genome increased significantly
from a scant 432 reads at 4 hpi to 16.9 million reads at 72 hpi
(\textbf{Table 1}, \textbf{Figure 2A}). From these mapped reads, we
identified 2,457 unique THEV splice junctions across all time-points,
with later time-points exhibiting significantly more sequence reads
supporting the splice junctions than earlier time-points. For instance,
all 13 unique junctions at 4 hpi had fewer than 10 supporting reads
each, averaging only 2.8 reads per junction. In contrast, the 2374
unique junctions at 72 hpi averaged 898.4 reads per junction, with some
junctions reaching as high as 322,677 reads. The marked increase in
splice junction and mapping reads to the THEV genome over time indicates
an active infection and successful viral replication, which is
corroborated by our quantitative PCR (qPCR) assay that quantified the
total number of viral genome copies over time (\textbf{Figure 2B}).

@@

The E2B region transcripts also start with the E2-5'UTR but extend
downstream to reach the TTS at 2334 bp in the IM region, which is in the
immediate context of a canonical PAS (position 2333 bp) where
polyadenylation likely occurs. The E2B transcripts, TRXPT\_6 and
TRXPT\_7, are almost identical except for an extra splice junction at
the 3'-end of TRXPT\_6 (\textbf{Figure 7}). TRXPT\_7 has the CP for both
classical proteins (pTP and Ad-pol) encoded in this region, with the pTP
ORF predicted to be spliced from two exons. The predicted splice
junction of pTP is corroborated by our data but the full transcript is
markedly longer than the predicted ORF, although the encoded product
(pTP) remains unchanged. Ad-pol (polymerase) is encoded downstream of
pTP with secondary SC (secSC) usage. The CP of TRXPT\_6 slightly differs
from TRXPT\_7 because a new STC resulting from the extra splice site
forms a minimal truncation of the Ad-pol encoded from its secSC.

@@

\textbf{MLTU or MLP Region transcripts}\\
The MLTU transcripts, dominant in the late phase of the AdV infectious
cycle, are produced by alternative polyadenylation and splicing of a
primary transcript and grouped into five transcript classes (L1-L5).
About 13 out of the 23 predicted ORFs in THEV fall within this TU, some
of which we have categorized under the E3 TU instead. Our RNA-seq data
revealed 12 transcripts (TRXPT\_8, TRXPT\_9, TRXPT\_10, TRXPT\_11,
TRXPT\_12, TRXPT\_13, TRXPT\_14, TRXPT\_16, TRXPT\_17, TRXPT\_18,
TRXPT\_19, TRXPT\_20) in this TU, most of which have the 5' TPL sequence
as in all AdVs. However, three transcripts (TRXPT\_16, TRXPT\_17,
TRXPT\_18) use a different leader sequence (sTPL), where a different
first exon is used instead of the first TPL exon, and TRXPT\_20 uses
only the third TPL exon (TPL3); see \textbf{Figure 10}.

@@

While the advent of next-generation sequencing has rendered easier the
study of large and complex eukaryotic transcriptomes, studying the
smaller, compact viral transcriptomes is counterintuitively more
challenging, as the transcripts typically have significant overlaps due
to genome economization. AdV transcriptomes escalate the difficulty due
to the wide array of mRNAs produced via very complex alternative
splicing and polyadenylation, all initiated from relatively few
promoters. Standard RNA-seq analysis programs, not primarily designed
for such compact, gene-dense, and complex transcriptomes, further
compound this challenge. Furthermore, in our case, no prior
transcriptomic studies for THEV exist; hence, assembling the transcripts
without any prior experimentally-derived annotation of THEV splicing
using only short illumina reads proved difficult. Lastly, we had
initially planned to sequence RNA from another time point (8 hpi);
however, all the RNA samples from 8 hpi and one replicate sample from 12
hpi got too degraded during the library preparation steps to be yield
any useful data. We believe that these would have contributed to better
insights into the temporal expression levels and splicing. Our approach
combines standard RNA-seq analysis programs with custom analyses and
experimentally validating all splice junctions with independent methods.
The transcript map for THEV produced from our analysis is strikingly
similar to that of the MAdVs.

Our work provides the first insights into THEV splicing, revealing 34
transcripts grouped into five TUs. The general temporal gene expression
regulation observed in MAdVs, with early regions peaking at earlier time
points followed by MLTU predominance at later time points, seems to also
apply to THEV. An unexpected observation is that the pileup of mapped
reads to THEV seems consistently skewed over similar regions of the
genome at all time points. Given the temporal regulation of AdVs gene
expression, we anticipated distinct differences in read pileups over the
genome at different time points, indicating the different stages of
infection. This may be due to an unsynchronized infection, leading to
transcripts overlapping the time points. Further research is required to
determine the precise temporal regulation of THEV.

Short read deep sequencing effectively reconstructs full AdV mRNA
structures, particularly mapping splice sites (18). However, the
substantial overlapping nature of AdV mRNAs and fragmentation during
library preparation make it challenging to map the exact TSS, TTS, and
PASs of assembled transcripts. As AdVs make heavy use of alternative
polyadenylation, short read RNA-seq is ill-equipped to discriminate mRNA
variants of the same gene produced via alternative polyadenylation.
Thus, shorter variants of alternatively polyadenylated mRNAs may
potentially be incorporated into the longer variants during transcript
assembly, significantly diminishing the diversity of mRNA in the
transcriptome. Also, independent transcripts with significant overlaps
may be assembled as a single, longer mRNA, since the short reads alone
do not provide enough context for the transcript assembler (StringTie)
to distinguish them. Such fusions may affect the transcript expression
level estimations by inflating or deflating the expression levels of the
transcripts involved, affecting the proper understanding of the temporal
gene expression regulation and also the diversity of the transcriptome.
Transcripts that have reads mistakenly fused with them would have
inflated expression levels while those whose reads are counted elsewhere
would show false lower expression levels. In our case, since we used
other independent methods to validate the splice junctions, we believe
these drawbacks to be minimized. Our results show transcripts in the
same TU initiating or terminating in similar areas, but not at the exact
same position. We consider the most upstream TSS or most downstream TTS
for for the transcripts involved but we present them unchanged in all
the figures shown (see \textbf{Supplementary Table S2}). Also, comparing
our results to the better-studied MAdV transcriptomes, we believe some
long transcripts in the MLTU (TRXPT\_9, TRXPT\_10, and TRXPT\_11) are
likely due to fusing some E3 transcripts to the terminal exons of the
MLTU transcripts by StringTie, making them significantly longer. These
mRNAs have unusually many exons and their last few exons are identical
to some E3 mRNAs. Future studies using long read sequencing technologies
will provide more precise mapping of the TSS, TTS, and PASs and clarify
the structures of the long MLTU transcripts.

While most predicted ORFs are encoded by the spliced transcripts, we
found some that seem to be truncated predictions, as either an upstream
in-frame SC or unknown upstream exons were found. Other ORFs were
identified that were either shorter or longer isoforms of some predicted
ORF. We also found several novel unpredicted ORFs (\textbf{Supplementary
Table S3}). On this basis, we anticipate that further studies will
likely reveal more unpredicted novel ORFs or new variants of predicted
ORFs. Furthermore, it is not unreasonable to presume that several splice
variants will likely be found as evidenced firstly by finding unique
transcripts using 3' RACE and during our splice junction validation
steps. And secondly, recent studies (17, 18, 22) are still discovering
novel mRNA variants for even the best studied MAdVs decades later. These
new potential proteins and isoforms significantly extend the repertoire
of THEV gene products than predicted, adding a hefty number of proteins
of undefined function to the previously predicted five (See
\textbf{Figure 1}). These new potential proteins and isoforms of unknown
functions may mediate or contribute to viral replication efficiency, or
the immunosuppression associated with THEV. Hence, further studies of
these potential proteins is urgently needed.

Eukaryotic mRNAs are typically functionally monocistronic, with the
5'-most AUG determining the translation reading frame. However, AdV
mRNAs, which span more than one ORF, are functionally polycistronic,
employing non-standard mechanisms of translation initiation such as
secSC usage and ribosome shunting (10, 22). AdVs use secondary AUGs as
initiation codons for most E1b proteins and some E3 proteins. In fact,
recent studies show that secSC usage is found transcriptome-wide. This
is thought to occur because translation initiation at the first SC is
inefficient, allowing downstream SCs to be employed (17). Ribosomal
shunting or jumping mechanism is utilized for MLTU transcripts that have
the TPL. This mechanism allows the ribosome to translocate to a
downstream AUG, under the direction of the shunting elements in the TPL,
even if a SC in a good Kozak sequence context is bypassed. Thus,
predicting the protein(s) expressed from an AdV mRNA is uncertain as any
one of the AUGs may be selected (10, 22). Almost all the THEV
transcripts in our data have the CP for several ORFs, some spanning as
many as six ORFs. This supports the usage of these special ribosome
initiation mechanisms as several predicted and novel ORFs found on mRNA
in our data could not be translated using only the typical ribosome
scanning mechanism. Interestingly, several distinct THEV mRNAs have
identical CPs. This is also observed in human AdVs in a recent study
(17). They proposed that this may permit protein production to be
fine-tuned through alteration in the balance between different mRNA
groups expressing that ORF.

@@

The THEV transcriptome bears remarkable similarity to the better-studied
MAdVs. The transcriptome is organized into five TUs, with temporal
regulation divided into early and late genes, and a broad repertoire of
transcripts are produced via virtually unlimited alternative splicing.
However, the THEV transcriptome appears less sophisticated (i.e, it
encodes fewer genes) than MAdVs, primarily because the MAdV genomes are
close to twice as long as that of THEV. The lack of subdivision of the
E1 region into E1a and E1b is one of the most obvious examples. Also,
the MAdV E4 region encodes several proteins unlike in THEV where only
one transcript encoding one protein was found. The complexity of the
MLTU leader sequences is another example. While the majority of the THEV
MLTU transcripts begin with the TPL just like MAdVs with a small subset
using a variant leader sequence (sTPL), significantly more diverse
5'-UTRs are employed for MAdV MLTU transcripts. Namely, the TPL, the
so-called x, y, and z leaders, and the i-leader are 5' leaders utilized
by MAdV MLTU mRNAs. The absence of these non-TPL leaders in our data
could mean that the 5'-UTR diversity of THEV's MLTU mRNAs is more
limited due to its smaller genome size or future studies could uncover
more variety not seen in our results. We also note that although THEV
genomic sequences show minimal differences between strains (12), the
transcriptomes may have significant variations; hence, our results may
vary from other THEV strains. Also, performing the study \emph{in vivo}
or with primary turkey cells may show different results. The potential
new proteins identified in our work adds to the number of proteins with
undefined functions in THEV; these may have roles in viral replication
efficiency, or the immunosuppression associated with THEV. Hence,
further studies of these proteins and other predicted proteins of
unknown function should be useful in elucidating THEV-induced
immunosuppression. Being the first transcriptomic characterization of
THEV, this work should serve as useful resource to future THEV gene
expression and transcriptomic studies, especially, mapping the mRNA
splice sites. More importantly, this work characterizing the splicing of
THEV mRNAs will allow researchers to accurately clone any THEV gene(s)
of interest to study its potential role in inducing immunosuppression or
other functions.
